---
title: "pRDA_outliers_detection"
author: "Thomas Francisco"
date: "2024-04-23"
output: html_document
---


```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(cache = FALSE)

library(vegan)
library(dplyr)
library(robust) #for rdadapt
library(qvalue) #for rdadapt
library(ggplot2)


```



################################ correcting for population structure #######################################

We will correct for population structure using only the PCa axis to avoir overcorrecting and to be consistent with other analysis (LFMM)


    1. DATA
    
They are coming from the script variance_partitioning

```{r data}

#genomic data
load("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic_new/data_allelic_frequencies_29pop_adapcon_gentree_475_8616.Rdata")

genomic_matrix <- data_allelic_frequencies_29pop_adapcon_gentree_475_8616

#climatic/structure/IBD
load("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/GEA/variance_partitioning/Data_RDA_pRDA_T.Rdata")


```

    2. Performing the RDA

We perform the pRDA accounting for population structure
```{r pRDA with population correction, message=FALSE, warning=FALSE}
#model
pRDA_env <- rda(formula = genomic_matrix ~  Annual_Tc+Tc_diur_range+Tc_Seasonality+Annual_P+P_Seasonality + Condition(PC1+PC2), data = Data_RDA_pRDA_T, scale=F)

RsquareAdj(pRDA_env)
summary(pRDA_env)
#significativity of the model
anova.cca(pRDA_env)

#test the significativity of the axis and the climatic variables using permutation
anova.cca(pRDA_env, step = 1000, by = "axis")
anova.cca(pRDA_env, step = 1000, by = "term")

```



Then, we need to perform the outliers selection on this models. To do this, we will follow procedure in Capblancq et al. 2021 and we will identified outliers based on their extremeness along a distribution of Mahalanobis distances estimated between each locus and the center of the RDA space using a certain number of axes (K). 

First, we need to select the number of ordination axis (RDA axis) that we will retains
```{r screeplot}
#screeplot
plot(pRDA_env$CCA$eig, option="screeplot")
#or like that
screeplot_pRDA<- screeplot(pRDA_env)

 #save
png("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA/pRDA/screeplot_RDA.png")
screeplot_RDA<- screeplot(pRDA_env)
dev.off()

#explained variance along each RDA axis
pRDA_env$CCA$eig

vif.cca(pRDA_env)

```
Interpretation: based on this, we will keep the 2 first RDA axis to perform the outliers detection

We need to represent the snp in the RDa space and the individuals
```{r population representation RDA space}
#score along the 2 first RDA axis
score_climatic_var <- as.data.frame(scores(pRDA_env, choices=c(1:2), display="bp"))
Score_population <- data.frame(pRDA_env$CCA$u[,c(1,2)])

#meta_data
meta_data_pop <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Populations/taxus_sample_29pop.csv",h=T,sep=";",dec=",")

#merge for country info
Score_population_bis <- rownames_to_column(Score_population,"Population")
score_with_country_info <- merge(Score_population_bis,meta_data_pop[,c(1,2)],"Population")
score_with_country_info$Country <- as.factor(score_with_country_info$Country)

#explained variance along each RDA axis
explained_variance <- data.frame(pRDA_env$CCA$eig)*100/sum(pRDA_env$CCA$eig) # variance of each axis*100 divided by the sum of all the explained variance for all axis

explained_variance_round <- round(explained_variance$pRDA_env.CCA.eig,digits=1)

group_palette <- c("Bosnia"="orangered3", "France"="gold2","Germany"= "darkorchid3", "Greece"="navyblue", "Italy"="turquoise2", "Norway"="green3", "Slovakia"="blue", "Slovenia"="red", "Spain"="black", "Sweden"="gray", "Switzerland"="orange", "UK"="darkgreen")


##Biplot avec les individus et les variables climatiques
biplot_populations <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = gray(0.80), size = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = gray(0.80), size = 0.6) +
  geom_point(data = score_with_country_info, aes(x = RDA1 * 3, y = RDA2 * 3, colour = Country), size = 2, alpha = 0.8) +
  geom_segment(data = score_climatic_var, aes(xend = RDA1, yend = RDA2, x = 0, y = 0), colour = "black", size = 0.15, linetype = 1, arrow = arrow(length = unit(0.02, "npc"))) +
  geom_text(data = score_climatic_var, aes(x=1.1*RDA1, y=1.1*RDA2, label = row.names(score_climatic_var)), size = 2.5)+
  xlab(paste0("RDA 1 (",explained_variance_round[1],"%)")) + 
  ylab(paste0("RDA 1 (",explained_variance_round[2],"%)")) +
  facet_wrap(~ "Adaptively enriched RDA space") +
  scale_color_manual(name = "Countries", values = group_palette, labels = levels(score_with_country_info$Country)) +
  theme_bw(base_size = 11) +
  theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size = 11)) +
  labs(color = "Country")

 print(biplot_populations)
 
 #save
png("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA/pRDA/biplot_pRDA_populations.png");print(biplot_populations);dev.off()

```
Interpretation: We can see that populations more in altitude/ more in the est are on the left along the first axis and populations in less altitute wih warmer temperature are on the right -> RDA1 seems to be a continentality/altitude gradient.
RDA2 separate the populations based on the temperature and precipitation. Seems to be an altitude gradient. 



    3. Identification of outliers

We can perform the genome scan procedure to calculate the Mahalanobis distance and then qvalues and pvalues associate to this distance to do an outliers selection

```{r genome scan function}
rdadapt<-function(rda,K)
{
  zscores<-rda$CCA$v[,1:as.numeric(K)]
  resscale <- apply(zscores, 2, scale)
  resmaha <- covRob(resscale, distance = TRUE, na.action= na.omit, estim="pairwiseGK")$dist
  lambda <- median(resmaha)/qchisq(0.5,df=K)
  reschi2test <- pchisq(resmaha/lambda,K,lower.tail=FALSE)
  qval <- qvalue(reschi2test)
  q.values_rdadapt<-qval$qvalues
  return(data.frame(p.values=reschi2test, q.values=q.values_rdadapt))
}

#Perform the fucntion to calculate the mahalanobis distance and then pvalues/ qvalues

genome_scan_pRDA <- rdadapt(pRDA_env,2)

#plot the distribution of pvalues 
hist(genome_scan_pRDA$p.values)

#save the histogram
 png(filename="C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA/pRDA/pvalues_distribution.png")

# a histogram we want to save
hist(genome_scan_pRDA$p.values)

# call this function to save the file 
dev.off()

```
    4. Selection of a threshold
    
Now, we need to select a threshold and then save the outlier
```{r selection of outliers}
# qvalues < 0.05
thres_qvalues <- 0.05

outliers_qvalues_5perc_pRDA <- data.frame(Loci = colnames(genomic_matrix)[which(genome_scan_pRDA$q.values<thres_qvalues)], p.value = genome_scan_pRDA$q.values[which(genome_scan_pRDA$q.values<thres_qvalues)], contig = unlist(lapply(strsplit(colnames(genomic_matrix)[which(genome_scan_pRDA$q.values<thres_qvalues)], split = "_"), function(x) x[1])))

outliers_qvalues_5perc_pRDA

#threshold pvalues < 0.05 with Bonferonni correction
thres_pvalues <- 0.05/length(genome_scan_pRDA$p.values)

outliers_pvalues <- data.frame(Loci = colnames(genomic_matrix)[which(genome_scan_pRDA$p.values<thres_pvalues)], p.value = genome_scan_pRDA$p.values[which(genome_scan_pRDA$p.values<thres_pvalues)], contig = unlist(lapply(strsplit(colnames(genomic_matrix)[which(genome_scan_pRDA$p.values<thres_pvalues)], split = "_"), function(x) x[1])))

outliers_pvalues

#top 1%
##add colnames
genome_scan_pRDA$snp_names <- colnames(genomic_matrix)

outliers_top1perc <- genome_scan_pRDA %>% 
  arrange(p.values) %>%
slice(1:(0.01*nrow(.))) %>%  
  as.data.frame()


```




We also perform the outlier detection based on the loadings of loci along the RDa axis to search for extreme loadings on each retained axis
```{r extremness along RDA axis}
load.rda <- scores(pRDA_env, choices=c(1:3), display="species")

hist(load.rda[,1], main="Loadings on RDA1")
hist(load.rda[,2], main="Loadings on RDA2")

```

```{r perform the detection }
#function to do it
outliers_function <- function(x,z){
  lims <- mean(x) + c(-1, 1) * z * sd(x)     # find loadings +/-z sd from mean loading     
  x[x < lims[1] | x > lims[2]]               # locus names in these tails
}

cand1 <- data.frame(outliers_function(load.rda[,1],3)) # 3
colnames(cand1) <- c("pvalues"); cand1$snp_names <- row.names(cand1)
cand2 <- data.frame(outliers_function(load.rda[,2],3)) # 51
colnames(cand2) <- c("pvalues"); cand2$snp_names <- row.names(cand2)

#names of the outliers
outliers_extremness <- rbind (cand1,cand2)


```


    5. Graphic representations of the outliers: 
    
```{r mahalanobis distance FDR 0.05 threshold}

score_loci <- as.data.frame(scores(pRDA_env, choices=c(1:2), display="species", scaling="none"))
score_loci_outliers <- data.frame(names = row.names(score_loci), score_loci)
score_loci_outliers$FDR5 <- "Neutral"
score_loci_outliers$FDR5[score_loci_outliers$names%in%outliers_qvalues_5perc_pRDA$Loci] <- "outliers"
score_loci_outliers$FDR5 <- factor(score_loci_outliers$FDR5, levels = c("Neutral", "outliers"))
score_loci_outliers <- score_loci_outliers[order(score_loci_outliers$FDR5),]

#Biplot avec les loci outliers et les variables climatiques
 biplot_outliers_pRDA<- ggplot() +
  geom_hline(yintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
  geom_vline(xintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
  geom_point(data = score_loci_outliers, aes(x=RDA1*15, y=RDA2*15,colour=FDR5), size = 1.4) +
  geom_segment(data = score_climatic_var, aes(xend=RDA1, yend=RDA2, x=0, y=0), colour="black", size=0.15, linetype=1, arrow=arrow(length = unit(0.02, "npc"))) +
  geom_text(data = score_climatic_var, aes(x=1.1*RDA1, y=1.1*RDA2, label = row.names(score_climatic_var)), size = 2.5, family = "Times") +
  xlab(paste0("RDA 1 (",explained_variance_round[1],"%)")) + 
  ylab(paste0("RDA 1 (",explained_variance_round[2],"%)")) +
  facet_wrap(~"RDA space: outliers FDR 5%") +
  guides(color=guide_legend(title="Locus type")) +
  scale_color_manual(values = c("gray90", "#F9A242FF")) +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))

 print(biplot_outliers_pRDA)
 
 #save
png("C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA/pRDA/biplot_outliers_pRDA_FDR_5perc.png");print(biplot_outliers_pRDA);dev.off()

```


```{r extremness along axis, 1% rank}
score_loci_outliers$rank1 <- "Neutral"
score_loci_outliers$rank1[score_loci_outliers$names%in%outliers_extremness$snp_names] <- "outliers"
score_loci_outliers$rank1 <- factor(score_loci_outliers$rank1, levels = c("Neutral", "outliers"))
score_loci_outliers <- score_loci_outliers[order(score_loci_outliers$rank1),]

#Biplot avec les loci outliers et les variables climatiques
ggplot() +
  geom_hline(yintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
  geom_vline(xintercept=0, linetype="dashed", color = gray(.80), size=0.6) +
  geom_point(data = score_loci_outliers, aes(x=RDA1*15, y=RDA2*15,colour=rank1), size = 1.4) +
  geom_segment(data = score_climatic_var, aes(xend=RDA1, yend=RDA2, x=0, y=0), colour="black", size=0.15, linetype=1, arrow=arrow(length = unit(0.02, "npc"))) +
  geom_text(data = score_climatic_var, aes(x=1.1*RDA1, y=1.1*RDA2, label = row.names(score_climatic_var)), size = 2.5, family = "Times") +
  xlab("RDA 1 (86.5%)") + ylab("RDA 2 (5.1%)") +
  facet_wrap(~"RDA space: outliers rank 1%") +
  guides(color=guide_legend(title="Locus type")) +
  scale_color_manual(values = c("gray90", "#F9A242FF")) +
  theme_bw(base_size = 11, base_family = "Times") +
  theme(panel.grid = element_blank(), plot.background = element_blank(), panel.background = element_blank(), strip.text = element_text(size=11))

```



    6. Overlapping between methods

We want to see if the outliers from extremness and the fdr 5% are similar: 
```{r overlapping of the methods}

colnames(outliers_qvalues_5perc_pRDA)= c( "snp_names","p.value","contig")
#overlapping extremeness 2.5 and top 1%
overlapping <- merge(outliers_extremness,outliers_qvalues_5perc_pRDA,"snp_names")
```
Interpretation: 135 out of the 136 outliers identified by extremeness are overlapping with the 243 of qvalues. 


We will use FDR 5% as a threshold to identified snp
We save the fdr 5% outliers: 

```{r save FDR 5% outliers}
outliers_pRDA_FDR5perc_T_Adapcon_gentree <- outliers_qvalues_5perc_pRDA[,-3]

write_xlsx(outliers_pRDA_FDR5perc_T_Adapcon_gentree,"C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA/outliers/outliers_pRDA_FDR5perc_T_Adapcon_gentree.xlsx")
save(outliers_pRDA_FDR5perc_T_Adapcon_gentree, file="C:/Users/tfrancisco/Documents/Thèse/Results/species/taxus/GEA/outliers/outliers_pRDA_FDR5perc_T_Adapcon_gentree.Rdata")
```



