---
title: "Population_structure_T_new_dataset_MQrank"
author: "Thomas Francisco"
date: "2024-03-14"
output: html_document
---


```{r packages}
library(LEA)
library(ggplot2)
library(pcadapt)
library(vegan)
```

In this script, we will study the population structure of Taxus baccata Adapcon-Gentree dataset of 8252 SNPs, 452 trees and 4.37% of NAs.
We will study the structure using PCA and we will check if the structure of some individuals/populations is not driven by the percentage of missing data. To do so, we will plot the size of the sampled points on the PCa depending on their Na percentage and the position of the points will depend of their genetic variants.


```{r metadata}

#load metadata
meta_data_vcf <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Samples/samples_taxus_baccata_adapcon_gentree.csv",h=T,sep=";",dec=",")

meta_data_pop <- read.csv("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/Populations/taxus_sample_29pop.csv",h=T,sep=";",dec=",")
```

      1. Individual-level


We first perform a screeplot analysis to see how much of the genetic variation is explained by the PCs axis. 
```{r  screeplot: number of PC selection and PCA}
#selection number of axis => screeplot
#initial parameters
load(file = "C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic_new/structure/Dataset_PCA_8252SNP_452IND.Rdata")
gen_data_new <- Dataset_PCA_8252SNP_452IND[,-1]; row.names(gen_data_new)<- Dataset_PCA_8252SNP_452IND$VCF_ID
data <- gen_data_new
k <- 10

#function
screeplot_genetic <- function(data,k){
  
  data_PCa <- data  %>% dplyr::select(-c("na_percentage_indiv")) %>% t() %>% data.frame()
#format for pcadapt
data_pcadapt <- read.pcadapt(data_PCa)

#perform the Pca
Pcadapt_results <-pcadapt(data_pcadapt,K=k,method = "mahalanobis")

# choose the number of axis
 # Create ggplot object
 screeplot_data <- data.frame(
    PC_axis = seq_along(Pcadapt_results$singular.values),
    Explained_variance = Pcadapt_results$singular.values^2
  )

  gg_screeplot <- ggplot(screeplot_data, aes(x = PC_axis, y = Explained_variance)) +
    geom_bar(stat = "identity") +
    labs(x = "PC axis", y = "Explained variance") +
    theme_minimal()
  
  # Display the screeplot
  gg_screeplot
  print(gg_screeplot)
  return(Pcadapt_results)
        }
Pcadapt_object <- screeplot_genetic(data,k)

#interpretation: we keep the two first axis

meta_data_vcf_452 <- meta_data_vcf[meta_data_vcf$VCF_ID %in% row.names(gen_data_new),]

#Initial parameters
data <-Pcadapt_object #dataframe coming from the return of the screeplot_genetic function 
names <- gen_data_new #initial dataframe
axis <- 2 #number of retained axis based on the screeplot
meta_data <- meta_data_vcf_452#meta data



#for loop to do it for multiple PC axis like 1-2 but also 1-3, 2-3 etc
 Pca_genetic <- function(data, names, num_axes, meta_data) {
   #we can create 2 loops that would perform the Pca_function for each combination of axis retained: 
  for (i in 1:(num_axes - 1)) { #here, the first loop will select axis between 1 and numb_axis-1 so basically, if we take 3 axis, i will take values of 1 then 2, so PCx will ether be 1 or 2.
    for (j in (i + 1):num_axes) { #here, the second loop within the first one will take the values of j from i +1 to num axis so if num axis=3, we will have j ranging from 2 to 3 when i=1 and j=3 when i =2
      #to summaries, the loops will perform the function for the combination of axis: 1-2,1-3 and 2-3 for 3 retained axis. 
      
      PCx <- i
      PCy <- j
      
      # Calculate explained variance for chosen PCs
      explained_variance <- data.frame(t(round((data$singular.values[PCx:PCy]^2) * 100, digits = 1)))
      colnames(explained_variance) <- c("PCx", "PCy")
      
      #store the scores along the retained PC axis in a dataframe
      score_Pca_imputed <- data.frame(data$scores[, c(PCx,PCy)], row.names (names))
      colnames(score_Pca_imputed) <- c(paste0("PC", PCx), paste0("PC", PCy), "VCF_ID")
      
      # add country, population information
      PCa_df_imputed <- merge(score_Pca_imputed, meta_data, "VCF_ID")
      
      # genetic PCA
      ggplot_representation <- ggplot() +
        geom_point(data = PCa_df_imputed, aes(PCa_df_imputed[,2],PCa_df_imputed[,3],size = names$na_percentage, color = Country)) +
        scale_colour_manual(name = "Country",
                            values = c("orangered3", "gold2", "darkorchid3", "navyblue", "turquoise2", "green3", "blue", "red", "black", "gray", "orange", "darkgreen")) +
        scale_size(name = "Missing data (%)", breaks = c(0, 5, 10, 14), labels = c("0", "5", "10", "15")) +
        xlab(paste0("PC", PCx, " ", "(", explained_variance$PCx, "%", ")")) + 
        ylab(paste0("PC", PCy, " ", "(", explained_variance$PCy, "%", ")"))
      
      print(ggplot_representation)
    }
  }
 }
 
#run the function 
Pca_genetic(data,names,axis,meta_data)
```
   2. Population-level

```{r Screeplot and PCA}

load(file = "C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic_new/data_allelic_frequencies_29pop_adapcon_gentree.Rdata")
gen_data_new <- data_allelic_frequencies_29pop_adapcon_gentree
data <- gen_data_new
k <- 10

#function
screeplot_genetic <- function(data,k){
  
#format for pcadapt

#perform the Pca
test <- rda(data, scale=T)
screeplot(test, type = "barplot", npcs=10, main="PCA Eigenvalues")

        }
Pcadapt_object <- screeplot_genetic(data,k)

#interpretation: we keep the two first axis

meta_data_vcf_452 <- meta_data_vcf[meta_data_vcf$Population %in% row.names(gen_data_new),]

#Initial parameters
data <-Pcadapt_object #dataframe coming from the return of the screeplot_genetic function 
names <- gen_data_new #initial dataframe
axis <- 2 #number of retained axis based on the screeplot
meta_data <- meta_data_vcf_452#meta data



#for loop to do it for multiple PC axis like 1-2 but also 1-3, 2-3 etc
 Pca_genetic <- function(data, names, num_axes, meta_data) {
   #we can create 2 loops that would perform the Pca_function for each combination of axis retained: 
  for (i in 1:(num_axes - 1)) { #here, the first loop will select axis between 1 and numb_axis-1 so basically, if we take 3 axis, i will take values of 1 then 2, so PCx will ether be 1 or 2.
    for (j in (i + 1):num_axes) { #here, the second loop within the first one will take the values of j from i +1 to num axis so if num axis=3, we will have j ranging from 2 to 3 when i=1 and j=3 when i =2
      #to summaries, the loops will perform the function for the combination of axis: 1-2,1-3 and 2-3 for 3 retained axis. 
      
      PCx <- i
      PCy <- j
      
      # Calculate explained variance for chosen PCs
      explained_variance <- data.frame(eigenvals(test))
      proportion_variance <- data.frame(t(explained_variance[c(PCx,PCy),] / sum(explained_variance) *100)) %>% round(digit=1)
      colnames(proportion_variance) <- c("PCx", "PCy")
      
      #store the scores along the retained PC axis in a dataframe
      store_score_rda <- scores(test, axes = c(PCx, PCy),display="sites" ) %>% data.frame(row.names(data))
      colnames(store_score_rda) <- c(paste0("PC", PCx), paste0("PC", PCy), "Population")
      
      # add country, population information
      PCa_df_imputed <- merge(store_score_rda, meta_data, "Population") %>% group_by(Population) %>% slice_head(n = 1) %>% ungroup() %>% as.data.frame()  #keep only the first row of each groups
      
      # genetic PCA
      ggplot_representation <- ggplot() +
        geom_point(data = PCa_df_imputed, aes(PCa_df_imputed[,2],PCa_df_imputed[,3],size=8, color = Country)) +
        scale_colour_manual(name = "Country",
                            values = c("orangered3", "gold2", "darkorchid3", "navyblue", "turquoise2", "green3", "blue", "red", "black", "gray", "orange", "darkgreen")) +
        xlab(paste0("PC", PCx, " ", "(", proportion_variance$PCx, "%", ")")) + 
        ylab(paste0("PC", PCy, " ", "(", proportion_variance$PCy, "%", ")"))
      
      print(ggplot_representation)
    }
  }
 }
 
#run the function 
Pca_genetic(data,names,axis,meta_data)



```


```{r SNMF admixture}

#we need to remove the na_percentage row
data_PCa_new <- gen_data_new  %>% dplyr::select(-c("na_percentage")) %>% t() %>% data.frame()

#we need to put the snp in columns and individuals in rows and put the snp in numeric
data_PCa_new_f <- data.frame(t(apply(data_PCa_new, 2, as.numeric)))

#we write the geno file 
write.geno(data_PCa_new_f,"C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic/data_PCa_new_f.geno")

#we load the geno file to check if everything is ok
geno_new <- data.frame(read.geno("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic/data_PCa_new_f.geno"))
View(geno_new)

#we run the admixture analysis

project_no_mac_no_impute_low_na <- snmf("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic/data_PCa_new_f.geno",K=1:10,entropy=T,repetitions=10,project="new")
plot(project_no_mac_no_impute_low_na, col = "blue", pch = 19, cex = 1.2)
project = load.snmfProject("C:/Users/tfrancisco/Documents/Thèse/Data/Espèces/Taxus_baccata/genetic/data_PCa_no_mac.snmfProject")

#plot the results
for(k in 2:10){
  
  my.colors<- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd","#7f7f7f","#bcbd22", "#c49c94", "#f7b6d2", "#dbdb8d")
  best = which.min(cross.entropy(project_no_mac_no_impute_low_na, K = k))
  barchart(project_no_mac_no_impute_low_na, K = k, run = best,
border = NA, space = 0,
col = my.colors,
xlab = "Individuals",
ylab = "Ancestry proportions",
main = "Ancestry matrix") -> bp
axis(1, at = 1:length(bp$order),
labels = bp$order, las=1,
cex.axis = .4)
qmatrix = LEA::Q(project_no_mac_no_impute_low_na, K = k,run=best)
qmatrix_ID_low_na=data.frame(row.names(gen_data_new),qmatrix);colnames(qmatrix_ID_low_na)=col_names <- c("VCF_ID",paste0("Group", 2:ncol(qmatrix_ID_low_na)-1))

mega_meta_data_tot_map <- merge(meta_data_vcf_451,meta_data_pop,"Population")

matrix_pop <- merge(qmatrix_ID_low_na, mega_meta_data_tot_map,"VCF_ID")

  mean_admixture_pop <- matrix_pop %>% 
   group_by(Population) %>% 
  summarize(across((Group1):(paste0("Group",k)), ~ mean(.)))
  
  data_tot_maps <- merge(meta_data_pop[,c("Population","Country","Longitude","Latitude")],mean_admixture_pop,"Population");colnames(data_tot_maps)=c("Pop","Country","Longitude","Latitude",paste0("Group", 5:ncol(data_tot_maps)-4));data_tot_maps$Longitude=as.numeric(data_tot_maps$Longitude);data_tot_maps$Latitude=as.numeric(data_tot_maps$Latitude)

groups <- paste0("Group",5:ncol(data_tot_maps)-4)#assign the number of group in the map_plot

  # Obtain the blank map of the region
admin <- ne_countries(scale = "medium", returnclass = "sf")

# Plot the map with scatterpie
map_pop <- ggplot() + 
  geom_sf(data = admin, fill = gray(0.92), size = 0) +#background of the map
  geom_scatterpie(
    data = data_tot_maps,
    aes(x = Longitude, y = Latitude),
    cols = groups, # Adjust columns accordingly
    size = 0,#size of the bold (=gras)
    alpha = 1 #transparence
  ) +
  scale_fill_manual(values=my.colors)+
  coord_sf(xlim = c(-10, 30), ylim = c(36, 65), expand = FALSE) +#extension of the map
  theme_minimal() +
  labs(title = "Genetic Structure for each Populations")

print(map_pop)
}
```